LSST Alerts: Science-Driven Options for Packet Contents and Pre-Filtering

### 2020-02-28 draft-alertcontents.tex
intermediate edits

% To add a short-form title:
% \title[Short title]{Title}
\title[Alerts Menu]{LSST Alerts: Science-Driven Options for Packets and their Distribution}

@@

\author{%
M.~L.~Graham and the Data Management System Science Team, with contributions from the Transients and Variable Stars Science Collaboration
}

@@

\setDocAbstract{%
\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
A review and discussion of the variety of options for alert packets and their distribution, such as latency timescale, packet contents, pre-stream filtering, and so forth.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2019-11-14}{Conception.}{Melissa Graham}
  \addtohist{0.1}{2020-01-09}{First draft ready for input from others.}{Melissa Graham}
}

@@

The purpose of this document is to consider the science motivation for, and impact of, various options for alert packet contents and their distribution timescale.
The contents and delivery of alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public, as they are exempt from any proprietary period, and publicly accessible because they are distributed to community brokers, and not solely available via the LSST Science Platform (LSP).
In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.

It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout.
This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.
Reducing the packet size or increasing (slightly) the latency could enable the stream to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.
The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).
Reducing the contents should only be done if that information is easily and quickly obtainable elsewhere.

{\bf Document Overview ---} \S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
\S~\ref{sec:latency} find that most alerts-related science goals would be satisfied with a $\sim$5 minute alert timescale, but that the quickly evolving fields of fast radio bursts and gravitational wave events may soon provide more robust motivation for a 60 second latency.
\S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).

@@

In this section the scientific impacts of various options to reduce the size of alert packets (in order to, e.g., enable wider alert distribution, facilitate broker processing) are considered, as are potential additions to the alert packet contents which would improve the user's ability to classify and prioritize alert sources for follow-up.
The main conclusions of the following discussion are:
\begin{itemize}
\item {\bf Removing Histories (\S~\ref{sssec:packets_remove_hist}) ---} Removing the {\tt DIASource} histories from the alert packets that are transmitted to brokers that request this modification could be scientifically beneficial, but a list of all {\tt DIASource} identifiers associated with the {\tt DIAObject} at the time of alert generation must be added to those packets. 
\item {\bf Removing Stamps (\S~\ref{sssec:packets_remove_stamps}) ---} Removing the image stamps from all alert packets could be scientifically beneficial, but stamps should still be made during alert production and made publicly retrievable with a latency similar to that of alert distribution.
\item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
\item {\bf Adding Object Association Information (\S~\ref{ssec:packets_add}) ---} To better enable extragalactic transient science with brokers, two new {\tt DIAObject} catalog elements should be computed and included in the alert packets: (1) the {\tt objectId} for the three {\tt Object} catalog galaxies with the lowest separation distance (based on the galaxy's 2D luminosity profile) from the {\tt DIAObject}, and (2) the separation distances for those three {\tt Objects}.
\item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
\end{itemize}

@@

\subsection{Additional {\tt Object} Association Information}\label{ssec:packets_add}

The contents of the alert packet as defined in \citeds{LSE-163} includes the following with respect to associations with {\tt Object} catalog from the most recent data release:
\begin{itemize}
\item {\tt nearbyObj} ({\tt unit64[6]}), the {\it "closest {\tt Objects} (3 stars and 3 galaxies) in Data Release database"}
\item {\tt nearbyObjDist} ({\tt float[6]}), the {\it "distances to {\tt nearbyObj}"} in arcseconds
\item {\tt nearbyObjLnP} ({\tt float[6]}), the {\it "natural log of the probability that the observed {\tt DIAObject} is the same as the nearby {\tt Object}"}
\end{itemize}
For the latter, there is a footnote that says {\it "This quantity will be computed by marginalizing over the product of position and proper motion error ellipses of the {\tt Object} and {\tt DIAObject}, assuming an appropriate prior"}.

The current definitions of {\tt nearbyObj}, {\tt nearbyObjDist}, and {\tt nearbyObjLnP} are not as useful as they could be for transients in host galaxies. 
For extragalactic transients, the three nearest galaxies are not always the three most likely host galaxies, and the distance in arcseconds matters less than a separation distance that accounts for the galaxy's luminosity profile.
Furthermore, the definition of {\tt nearbyObjLnP} is only appropriate for static variable point sources (stars): for transients in host galaxies, the observed {\tt DIAObject} will never be ``the same as the nearby {\tt Object}".

Statistically, the most likely host for a given transient is the galaxy which contributes the most optical flux at the transient's location.
This is usually estimated by calculating an offset distance from the nearby galaxies to the transient that is expressed in terms of the galaxy's spatial luminosity profile, and assuming the galaxy with the lowest offset distance is the host.
The following are several options for estimating which nearby galaxy is the most likely host of an extragalactic transient.

{\bf Effective Radius --} Calculate a separation distance that is the radial distance from the core of the galaxy to the location of the transient, divided by the effective radius of the galaxy (i.e., {\tt kronRad90} in the {\tt Objects} table; \citeds{LSE-163}).
The nearby galaxy with the lowest separation distance is the most likely host. This will account for the relative sizes of the potential host galaxies, but not their position angles.

{\bf Second Moment --} Calculate a separation distance, as in \citet{2006ApJ...648..868S}, based on the two-dimensional luminosity profile of the galaxy.
Where $x_{\rm trans},y_{\rm trans}$ is the location of the transient, and $x_{\rm gal},y_{\rm gal}$ is the center of the galaxy (the first moment; {\tt radec} in the {\tt Object} table), the separation distance is $R^2 = C_{xx} x_r^2 + C_{yy} y_r^2 + C_{xy} x_r y_r$, where $x_r = x_{\rm SN} - x_{\rm gal}$ and $y_r = y_{\rm SN} - y_{\rm gal}$.
The ellipse parameters $C_{xx}$, $C_{yy}$, and $C_{xy}$ can be calculated directly from the second moments of the galaxy luminosity profile ({\tt Ixx}, {\tt Iyy}, and {\tt Ixy} in the {\tt Object} table), e.g., as described in Section 10 of E. Bertin's Source Extractor manual\footnote{Version 2.3: \url{https://www.astromatic.net/pubsvn/software/sextractor/trunk/doc/sextractor.pdf}}.
The nearby galaxy with the lowest separation distance is the most likely host.
This option accounts for both the relative sizes and position angles of the potential host galaxies, and requires only a little more processing.

{\bf 2D Algorithms --} There are also other, more complicated methods for identifying the most likely host for a given transient.
For example, the nearby galaxy with the smallest fraction of light interior to an isophot through the transient's location, where the isophot shape is given more degrees of freedom and not constrained to concentric ellipticals as in the second moment method above.
Another example is to use an algorithm that provides deblended footprints for nearby extended objects, and can estimate the fraction of light in given pixel that should be attributed to each (e.g., the SCARLET deblender, \cite{2018A&C....24..129M}).
\textcolor{red}{\citeds{LSE-163} does mention that footprints will be generated for {\tt Objects}, but it doesn't look like they'd be stored in the {\tt Object} table; find out if these footprints would be appropriate/available to use here.}
The most likely host galaxy would be the one which contributes the most flux at the pixel location of a transient.

Generally, nearby galaxies with separation distances $>$3, or $>$99\% of the luminosity profile, are not assigned as hosts and the transient is considered "hostless".
While this cutoff has been appropriate for past samples of $\sim$hundreds of transients, it will not be appropriate for the LSST sample size, and no such cut should be applied when evaluating potential host galaxies for {\tt DIAObjects}.

All of the above host galaxy identification methods can also make use of priors if the transient type is known; for example, the established correlation between core-collapse supernovae and star formation.
However, such a robust level of host associations are beyond scope for the Prompt pipeline, and the main goal is to provide brokers with sufficient information to assess the potential host association, and decide whether it is useful to the classification and/or prioritization of alerts for follow-up.

{\bf Summary --} For the ten nearest {\tt Object} catalog galaxies, a separation distance should be calculated with respect to the transient location, preferably using the second moments of each galaxy's luminosity profile. Two new {\tt DIAObject} catalog elements should be added: {\tt nearbyPotHost}, containing the {\tt objectId} for the three galaxies with the lowest separation distances, and {\tt nearbyPotHostSepDist}, containing the separation distances. An analog for the existing element {\tt nearbyObjLnP}, representing the probability, is not necessary. This would add {\tt unit64[3]} and {\tt float[3]} to the {\tt DIAObject} catalog and to each alert, but this both tiny and worthwhile.

@@

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % %
\section{Alert Distribution Latency} \label{sec:latency}

This section considers the scientific motivation for the 1 minute alert distribution latency, and the impact to alerts-related science if that latency was extended to, e.g., 5 minutes.
It is a requirement that the DMS be capable of supporting the distribution of at least 98\%\reqparam{OTT1}\reqparam{OTR1}\lsrreq{0101}\lsrreq{0025}\ossreq{0127}\dmreq{0004} of alerts for each visit within 60 seconds of the end of image readout.
However, if it is possible to relax this requirement without a negative impact on science then it might be possible for LSST to transmit alerts to $>$5 brokers (and/or repurpose human and computational resources dedicated to the 1 minute latency), which in turn might enable more alerts-related science.

Generally, for science goals that require alerts on timescales shorter than $\sim$5 minutes, this need stems from targets that are fast-evolving or short-lived (or both) and require follow-up on timescales shorter than $\sim$15--30 minutes.
Since most observing strategies for the WFD main survey do not revisit fields on timescales shorter than $\sim$15 minutes (strategies that do pair visits aim for closer to $\sim$22 minute gaps), brokers must be able to confidently identify such targets with single-epoch single-filter LSST photometry, or be using additional information (e.g., host galaxy characteristics, coincident alerts from other surveys).
For the purposes of this discussion, an optimistic outlook is assumed, and the difficulty in identifying targets with a single epoch of LSST photometry is not considered as a reason to downgrade the importance of a 1 minute alert latency.

After a consideration of the following science cases that are most likely to require rapid follow-up (summary below), it appears that most of them would be satisfied without risk if the latency was increased from 1 to 5 minutes. 
However, it is absolutely not unambiguously clear that a need for 1 minute alerts will not evolve out of the still emergent fields of fast radio bursts and gravitational wave events over the next couple of years.
\begin{itemize}
\item {\bf A TVS Survey (\S~\ref{ssec:latency_tvs})} found that only $<$10\% of respondents reported that requiring alerts within 1 minute was necessary for their science goals.
\item {\bf GW Follow-up (\S~\ref{ssec:latency_emgw}) ---} Although a 5 minute latency would suffice for kilonovae (days-long optical counterparts to neutron star mergers), some theoretical models predict optical emission within the first hour (e.g., jet break-out or spin-down energy from the merger). However, these GW scientific motivators for a 1 minute alert latency are moot if the time between GW event and LSST ToO follow-up remains at tens of minutes or more.
\item {\bf Gamma-Ray Bursts (\S~\ref{ssec:latency_grb}) ---} \textcolor{red}{Tentatively,} alerts for the coincident optical counterparts of short GRBs would be most scientifically useful if distributed within 1 minute, but the predicted occurrence rate is $\lesssim$20 over the 10-year LSST survey.
\item {\bf Fast Radio Bursts (\S~\ref{ssec:latency_frb}) ---} \textcolor{red}{Tentatively,} some models predict that a short, bright burst of optical emission from the physical mechanism powering FRBs could be detected by LSST 5 to 15 minutes prior to the radio signal.
If such optical events occur with significant frequency, are bright and consistently detectable, and can be identified with a single-filter detection, and if radio telescopes can respond within minutes, then LSST is the most promising future facility for this science, which requires a 1 minute alert distribution latency.
\item {\bf Young Supernovae (\S~\ref{ssec:latency_ysne}) ---} The science goals require follow-up in an hour or longer, and a 5 minute latency would not have a negatively impact.
\item {\bf Solar System Objects (\S~\ref{ssec:latency_sso}) ---} \textcolor{red}{TBD.}
\item {\bf Stellar Variables (\S~\ref{ssec:latency_stars}) ---} \textcolor{red}{TBD.}
\end{itemize}

\subsection{TVS Survey Results Regarding Alerts Timescales}\label{ssec:latency_tvs}
{\it Contents contain the results of a TVS survey administered by Rachel Street.}

The Transients and Variable Stars (TVS) science collaboration surveyed the science needs of their members with respect to alert latency, and the results\footnote{Results courtesy of TVS co-chair Rachel Street.} are shown in Table \ref{tab:tvs}.
The survey asked respondents to choose the maximum tolerable and ideal average delay between the alerts being produced by the LSST data reduction pipeline and the alert information becoming available through the broker service.
This is not exactly the same as the alert distribution timescale {\tt OTT1}, but these responses will inform the need for $<$1 minute alert distribution.
Respondents were asked to provide a short summary of their science goals for alerts if they reported needed access within 15 minutes.
Note that the pool of respondents is probably not representative of the wider collaboration, and is likely biased towards individuals with science interests that do require faster access to alerts.

\begin{table}[h]%[htdp]
\caption{Table of results from a TVS survey which asked "how fast do you really need alerts?". The total number of respondents was 20. The science driver acronyms are: EM-GW (electromagnetic counterparts to gravitational wave events), YSNe (young supernovae, including e.g., shock breakouts), GRBs (gamma-ray bursts). \label{tab:tvs}}
\begin{center}
\begin{tabular}{|l|cl|cl|}
\hline
             & Maximum & Science & Ideal       & Science \\
Latency & Tolerable  & Driver(s) &  Average & Driver(s) \\
\hline
1 min or less & 1 & EM-GW, GRB  & 2 & EM-GW, YSNe, GRB \\
1-5 min         & 0 &                          & 1 &                                   \\
5-10 min       & 1 & EM-GW, YSNe & 1 & YSNe \\
10-30 min     & 2 & YSNe               & 2 & EM-GW, GRBs \\
30-60 min     & 2 &                          & 7 & EM-GW  \\
1-6 hours     & 4 & EM-GW, GRBs & 1 &  \\
6-12 hours   & 2 &                          & 1 &  \\
12-24 hours & 3 & EM-GW            & 1 & EM-GW \\
1-3 days      & 1 &                           & 0 &  \\
>3 days       & 4 &                           & 4 &  \\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%

Only 20\% (10\%) of respondents report that $<$10 minutes ($\leq$1 minute) is an ideal average delay, and whereas 70\% report that $>$30 minutes would be a sufficient average latency.
The three science drivers associated with $<$5 minute alert access are the electromagnetic counterparts to gravitational wave events (EM-GW; kilonovae), young supernovae (early short-lived light curve features such as shock breakouts), and gamma-ray bursts (GRB).
The fact that these three are also listed as science drivers for longer-latency alert is mainly due to the diversity within the science drivers and the fact that the events have both short- and longer-timescale features.
Each of these science drivers is discussed in turn below, along with two other science cases that rely on rapid access to LSST alerts: fast radio bursts (FRBs) and solar system objects (SSOs).

\subsection{EM Counterparts to GW Events}\label{ssec:latency_emgw}
{\it Some contents revised from a Slack conversation between Federica Bianco, Om Sharan Salafia, and Eric Bellm.}

During LSST Operations, target-of-opportunity imaging follow-up sequences might be executed in the error ellipse of a gravitational wave (GW) detection to search for the electromagnetic (EM) optical counterpart.
Such a search yielded a fast-evolving "kilonova" which decayed from 22 to 28th magnitude in the $g$-band in just 6 days \citep[faster in the bluer and slower in the redder filters][]{2017Sci...358.1559K}.
Although the kilonova light curves last only a few days, since there is so far only one event, GW170817, a day or two delay between the GW event and the optical detection still yields very scientifically valuable data -- for now.
During LSST Operations there will already exist a sizable collection of longer-latency follow-up, and it is likely that science will be moving in the direction of pushing to ever earlier detections.
However, since kilonovae produce days-long optical afterglows, a 5 minute LSST alerts would likely suffice.

There are two theoretical predictions for prompt optical emission that would require very rapid access to alerts.
One of them is a potential faint, $<$1 hour, UV/optical transient that occurs at the time of jet-break out for a short gamma-ray burst associated with a binary neutron star merger (\S~\ref{ssec:latency_grb}).
Another that predicts emission of a similar color, luminosity, and timescale is the spin-down energy of a long-lived ($10^2$--$10^4$ s) neutron star formed from a binary neutron star merger before its eventual collapse to a black hole \citep{2016ApJ...819...15S}.
As described in \S~\ref{ssec:latency_grb}, very few events could be detected by serendipitous coincidence by the LSST WFD main survey, and targeted follow-up of well-localized GW events with more appropriate facilities such as space-based UV/optical imagers is much more likely to yield detections of this very short lived emission.

There are two additional issues related to ToO for EMGW events which might make 1 minute alerts necessary or impossible.

First, for rapid ToO observations when visit coordinates are not known in advance, there might not be time to pre-cache the slow database queries that are needed during alert production processing (e.g., loading slices of the PPDB catalogs).
This might mean that an alert distribution latency cannot be guaranteed for ToO imaging surveys. 
For the purposes of this science-driven discussion, assume this to be a surmountable technical implementation issue.

The second is that, at least for run O3, the GW event detection system itself issues preliminary alerts within 1--10 minutes, and these preliminary alerts are often retracted and do not always have the sky localization.
Even now, many imaging follow-up surveys wait for the initial alert (or retraction) to be sent after a round of human vetting of the GW event signal, and this can take several hours.
With such latencies, the question of 1 {\it vs.} 5 minute LSST alert timescales becomes inconsequential.

\subsection{Gamma-Ray Bursts}\label{ssec:latency_grb}

{\it \textcolor{red}{MLG: no outside input on this yet, is needed to be confirmed by a GRB person.} }

Short gamma-ray bursts (sGRBs) are thought to be the mergers of two neutron stars, or a neutron star and a black hole.
As the jet propagates through the ejecta material a hot cocoon is formed.
When cocoon and jet break out, along with the sGRB can be observed a blue optical transient with an absolute peak brightness of -12 to -15 magnitudes which lasts for $10^3$--$10^4$ seconds (e.g., \citealt{2018MNRAS.473..576G}).
For an LSST detection limit of $r\sim24$ mag, a detection limit of $r\lesssim-12$ ($r\lesssim-15$) mag corresponds to distances of $\lesssim160$ ($\lesssim630$) Mpc, or redshifts $z\lesssim0.035$ ($z\lesssim0.14$).
Observing the diversity of cocoon emission for a sample of sGRBs would require regular and consistent multi-band follow-up within $10^3$ seconds, or $15$ minutes.
In such cases where a sGRB is observed in a time/area that is serendipitously coincident LSST visit, science would be maximized if the alert confirming the presence of an optical counterpart was distributed to brokers within 1 minute.
However, what is the expected frequency of this situation occurring during LSST Operations?

% z=0.035 is V=0.014 Gpc3
% z=0.14 is V=0.836 Gpc3
% (0.836/0.014) * 4/yr = 240/yr
The rate of short GRBs in the local volume ($<$200 Mpc) is estimated to be quite low, $<$4 $\rm yr^{-1}$ \citep{2019arXiv190800100M}.
Scaled up to the co-moving volume within 630 Mpc, this rate is $\sim$240 $\rm yr^{-1}$, all-sky.
In the baseline main survey about a sixth of the sky can be observed in a given night, which implies $\sim$1 sGRBs with detectable optical counterparts in the WFD survey per night for every $\sim$10 nights of observations.
The probability that one of the 1000 visits per night will occur during the 5 minutes in which an alert would be useful for triggering rapid follow-up is $<10^{-2}$.
Very roughly, we might expect this serendipitous coincidence to occur twice a year at most, or $\sim$20 times over the 10-year LSST main survey.

% 10 visits per 5 minute block, or 100 "blocks" per night
% 100x100 potential sky area/time options for the 100 blocks, x10 days
% of the options, 100 blocks per night x 10 nights would be done
% and then randomly draw just the 1 sGRB
% scipy.stats.hypergeom(100000,1000,1)
% 0 0.9900 
% 1 0.0100 
% 2 0.0000

For discovering the optical counterparts of sGRBs, the LSST wide-fast-deep survey is simply not the most efficient future facility for the job.
As described by \citet{2018MNRAS.473..576G}, a rapid search at the location of sGRBs with a UV satellite such as ULTRASAT would be ideal.

%Triggering criteria might be:
%new {\it u}- or {\it g}-band source with recent limits associated with a $<$200 Mpc galaxy (how many expected per night?)
%and then very high priority if it's serendipidously associated with a GRB detection in the last 5 minutes

\subsection{Fast Radio Bursts}\label{ssec:latency_frb}

{\it Some contents based on text from and discussions with Tyler Pritchard.}

One of the most likely -- but also most mysterious -- transients that might significantly benefit from the $60$ second alert timescale are fast radio bursts (FRBs): a millisecond long pulse of coherent emission in the GHz range.
The emission is dispersed by the inter-galactic medium (IGM), such that the pulse's observed arrival time is frequency dependent.
Observed FRB dispersion measures of $\rm{DM}\approx100$ to $1000$ $\rm pc\ cm^{-3}$ indicate that they originate at cosmological distances, with redshifts $z\approx0.1$ to $1$ \citep{2018Natur.562..386S}.

If an optical counterpart is generated by this coherent emission, the time delay between the optical detection and the radio detection is on the timescale of minutes for frequencies $\nu < 500\ {\rm MHz}$, as shown in Figure \ref{fig:sci_frb}.
This leaves open the possibility for triggering radio follow-up of an optical counterpart candidate serendipitously detected by LSST -- if such counterparts exist and are detectable.
A millisecond-long event in the optical would have to be quite bright to be detected by LSST, but studies show that LSST detections are feasible \citep{2016ApJ...824L..18L}.
\cite{2019ApJ...878...89Y} demonstrate that two theoretical sources for coherent optical emission from FRBs would not be detectable by LSST, but that inverse Compton scattering processes could lead to optical detections with LSST (e.g., from pulsars or masers).
Searches for FRB counterparts in optical images that were serendipitously obtained (near-) coincidentally with an FRB detection have only just recently become possible thanks to current wide-field imaging surveys.
No transient optical counterparts have yet been detected \citep{2019ApJ...881...30T}, but individual host galaxies for FRB events have been identified \citep{2016Natur.530..453K}.

Since the rate of FRBs is estimated to be quite high (thousands per day; \citealt{2016MNRAS.460L..30C}) even after radio detection efficiency is factored in there could be many within the observed LSST volume. 
For example, if there are 1000 FRB occurring across the full sky every 24 hours, and LSST observes one sixth of the sky during a night's 10 hours of darkness, then there would be $\sim70$ FRB occurring in that area during that time.
Assuming that both the FRB and the LSST visits are distributed randomly in time and location (although the visits are correlated, obviously), there's a $\sim6\%$ chance that in any given night one visit will be serendipitously coincident with the expected arrival of a short lived optical precursor to the FRB.
If such a source could be recognized as an FRB optical counterpart within a minute, and radio follow-up could begin as soon as possible to detect the radio component and measure the time delay (Figure \ref{fig:sci_frb}).  
However, \textcolor{red}{it is (or rather, MLG is) currently unclear whether (a) there is strong theoretical support for such optical counterparts, and (b) radio telescopes can perform ToO follow-up on $<$15 minute timescales}.

If FRBs are associated with superluminous supernovae (SLSNe) and young magnetars, then potentially an LSST alert regarding a change in behavior of known SLSNe could be used to trigger radio follow-up.
However, this case is unlikely to be as sensitive to the LSST alert distribution timescale as the optical emission would be released over a longer time window \citep{2019arXiv191002036L}.

\begin{figure}[h]
\begin{center}
\includegraphics[width=15cm]{figures/frb_optical_delays.png}
\caption{The delay time (for coherent emission) between the arrival of the optical and radio photons due to cosmological dispersion, as a function of the frequency of the radio detector. The {\it lowest} frequency bands of the Square Kilometer Array (SKA) and the Low-Frequency Array (LOFAR) are marked with vertical dotted and dash-dotted lines, and the 1 minute timescale for LSST alert distribution with a horizontal dashed line. The delay time is $\Delta t = k_{\rm DM}\,DM\,(\nu_{\rm low}^{-2} - \nu_{\rm high}^{-2})$, where $k_{DM}=4.149$ $\rm GHz^2\ pc^{-1}\ cm^3\ ms$, ${\rm DM}$ is the dispersion measure in $\rm pc\ cm^{-3}$, $\nu$ is the low and high frequency bands of the observation, and $\Delta t$ is in $\rm ms$. We have used the range of FRB dispersion measures as observed by \cite{2018Natur.562..386S}. \label{fig:sci_frb}}
\end{center}
\end{figure}

\subsection{Young Supernovae}\label{ssec:latency_ysne}

Photometric data obtained during the first few hours to days of a supernova can constrain the progenitor radius and circumstellar material in the immediate environment, which in turn provides information about the progenitor star, its system, and its final stages of evolution before the explosion.
For example, \citet{2018Natur.554..497B} model the optical photometric data of the post-shock breakout cooling peak for a massive star's core collapse supernova to constrain the radius and mass of its outer envelope; \citet{2012ApJ...744L..17B} used the first day of photometric observations of a nearby Type Ia (thermonuclear detonation) supernova to show that their progenitor stars must be a compact degenerates; and rare ``blue bumps" in the first few days of a Type Ia supernova's light curve might betray the presence of a non-degenerate companion star \citep[e.g.,][]{2010ApJ...708.1025K,2017ApJ...845L..11H}.

For young supernovae, follow-up observations within a couple of hours are required, and the related science goals are unlikely to be impacted if the alert latency was 5--10 minutes instead of 1 minute.

\subsection{Solar System Objects}\label{ssec:latency_sso}

\textcolor{red}{MLG: in Lynne's talk on the SSSC needs for alerts brokers at the CBW in June 2019, slide 4 says that a needed broker capability was "API or other 'push' notification (e.g., trailed detections) [minutes may count]". Follow-up on the SS use case for alerts on very short timescales.}

\subsection{Stellar Variables}\label{ssec:latency_stars}

\textcolor{red}{Return to the road map or science book? Surely there must be some stellar use-cases for 1 minute alerts. E.g., microlensing peaks can be real sharp.}

>>>>

% To add a short-form title:
% \title[Short title]{Title}
\title[Alerts Contents]{LSST Alerts: Science-Driven Options for Packet Contents and Pre-Filtering}

@@

\author{%
M.~L.~Graham and the Data Management System Science Team
}

@@

\setDocAbstract{%
\textcolor{red}{\bf EXTREME DRAFT STATE. DO NOT CITE. DO NOT READ. LOOK AWAY.}\\
A review and discussion of the variety of options for alert packets contents, pre-stream filtering, and so forth.
}

% Change history defined here.
% Order: oldest first.
% Fields: VERSION, DATE, DESCRIPTION, OWNER NAME.
% See LPM-51 for version number policy.
\setDocChangeRecord{%
  \addtohist{0}{2019-11-14}{Inception.}{Melissa Graham}
  \addtohist{0.1}{2020-01-09}{First draft.}{Melissa Graham}
}

@@

The alert packets will play a major role in the scientific impact of LSST, in part because they are the only LSST data product that is both world public (exempt from any proprietary period) and publicly accessible, as they are distributed to community brokers and not solely available via the LSST Science Platform (LSP).
In comparison, although the contents of the Prompt products database (PPDB) from which alerts are generated are also world public, access to the PPDB is restricted to individuals with data rights \citedsp{LDO-013} who have authenticated accounts in the LSP.

This document considers options for the alert packet contents and potential ``pre-filters" of the alert stream based on those contents. 
The alert packets, as described in \citeds{LSE-163}, were designed to include all the relevant LSST data about a source that a broker might need to assess, classify, and prioritize the alert for follow-up observations \emph{within minutes} (i.e., on order the timescale of the 60 second alert distribution latency).

It is a requirement\reqparam{numStreams}\dmreq{0391} that the DMS be capable of supporting the transmission of at least 5 full alert streams within 60 seconds of image readout.
This is based in part on estimates of the alert stream data rate and the bandwidth allocated to alert distribution in the LSST data facility.

Reducing the packet size by narrowing down the contents included, and/or reducing the number of alerts distributed by applying a pre-filter, could enable alerts to be delivered to more brokers, and thereby increase the amount of alerts-related science done by the community.

\S~\ref{sec:packets} discusses how removing stamps and histories from alerts could be scientifically beneficial for some brokers, as long as stamps are made publicly available with similar latency, and proposes additional alert elements to better identify potential transient host galaxies.
\S~\ref{sec:prefilter} lists types of pre-filtered streams that could improve brokers' scientific productivity and estimates the resulting bandwidth reductions, and \S~\ref{sec:graceful} argues that when alerts are delayed, the scientifically optimal response would be to flag and distribute them as soon as possible (and not, e.g., send them directly to an archive).

@@

In this section the scientific impacts of various options to reduce the size of alert packets (in order to, e.g., enable wider alert distribution, facilitate broker processing) are considered, as are potential additions to the alert packet contents which would improve the user's ability to classify and prioritize alert sources for follow-up.
The main conclusions of the following discussion are:
\begin{itemize}
\item {\bf Removing Histories (\S~\ref{sssec:packets_remove_hist}) ---} Removing the {\tt DIASource} histories from the alert packets that are transmitted to brokers that request this modification could be scientifically beneficial, but a list of all {\tt DIASource} identifiers associated with the {\tt DIAObject} at the time of alert generation must be added to those packets. 
\item {\bf Removing Stamps (\S~\ref{sssec:packets_remove_stamps}) ---} Removing the image stamps from all alert packets could be scientifically beneficial, but stamps should still be made during alert production and made publicly retrievable with a latency similar to that of alert distribution.
\item {\bf Distributing Multiple Packet Formats (\S~\ref{sssec:packets_remove_procon}) ---} From a science perspective, distributing non-identical alert packets would have no negative impact provided that full-sized packets do not become more prone to delays.
\item {\bf Compressing Packets (\S~\ref{ssec:packets_compress}) ---} From a science perspective, compressing the alert packets has little scientific impact unless it imposes a large additional latency on alert distribution, which could be detrimental to alerts-related science.
\end{itemize}

@@

@@
